---
title: "Linear generative models"
author: Deven Gokhale
format: 
  html:
    theme: sketchy
    code-fold: true
editor: visual
---

```{r, setup, warning=FALSE, message=FALSE}
rm(list = ls())

# additional packages used for bayesian analysis of the model
library(brms)
library(bayesplot)
library(tidybayes)
library(GGally)
library(psych)
library(tidyr)
library(dplyr)

theme_set(theme_bw())
# setting up some stan options to make it run faster 
# Load necessary libraries and set up multi-core processing for Stan
options(warn=-1, message =-1)

#rstan_options(auto_write = TRUE)

# assign multiple cores to the estimation values 
#options(mc.cores = detectCores()-4)


# extract data from the rethinking pacakge.
data(Howell1, package = "rethinking")

hd <- Howell1

rm(Howell1)

hd |> describe()

# how do the distibutions of the continuous variables look
hd |> 
    select(height, weight) |> 
    pivot_longer(cols = c(height, weight)) |> 
    ggplot(aes(x = value))+
    geom_histogram(bins = 50)+
    facet_wrap(name~., scales = "free_x")



hd |> 
    mutate(adult = ifelse(age >= 18, 'yes', 'no')) |> 
    ggplot(aes(x = weight, y = height, colour = adult))+
    geom_point(pch = 21, fill = NA)
```

When stratified by being adult, the relationship shows a strong linear relationship between height and and weight. For the remainder of this workflow, only I will only be using the adult data since it more applicable to this analysis.

### Linear regression

The main aim of this portion is to introduce a simple linear regression model and run it using the `brms` package. I will then proceed to extract useful summary statistics from the fitted model. Basically trying to get used to the processing and post-processing workflows with these newer packages.

```{r, fitting_linear_model}

# subset the Howell data to include only adults 
hd_adults <- hd |> filter(age >= 18) 

# a quick plot to visualize the distribution of the height and weights
hd_adults |> 
  select(height, weight) |> 
  pivot_longer(cols = c(height, weight)) |> 
  ggplot(aes(x = value))+
  geom_histogram(bins = 50)+
  facet_wrap(name~., scales = "free_x")

# pretty unimodal once filtering based on age. cool! 
    
# a quick plot to look at the relationship between height and weight
hd_adults |> 
  ggplot(aes(x = height, y = weight))+
  geom_point(pch = 21, fill = NA, )

# perfecto! a strong linear relationship exists between height and weight!


# summarizing the values for the two variables for adulst 
hd |> 
  mutate(adults = ifelse(age >= 18, 'yes', 'no')) |> 
  select(adults, age, height, weight) |>
  group_by(adults) |> 
  summarise_all(
    .funs = list(Mean = mean, SD = sd, Median = median)
  ) |> 
  ungroup()

hd |> 
  mutate(adults = ifelse(age >= 18, 'yes', 'no')) |> 
  group_by(adults) |> 
  tally() |> 
  ungroup()
```

Taking a second to first generate a prior probability model for the the height distribution. Assuming the empirical statistics to generate the following formalism. Say the height $h_i \sim N(\mu, \sigma)$ where $\mu \sim U(154, 7)$, and $\sigma \sim U(0, 50)$

```{r, height_prob_model, message=FALSE, warning=FALSE}

set.seed(72)

prior_pred_model_height <- (
    tibble(
        mu_draws = rnorm(1e4, mean = 178, sd = 100), 
        sigma_draws = runif(1e4, 0, 50),
        ) |> 
    mutate(
        height_draws = rnorm(1e4, mean = mu_draws, sd = sigma_draws)
        )
)

# plot the prior distribution of height

prior_pred_model_height |> 
    pivot_longer(cols = mu_draws:height_draws) |> 
    ggplot(aes(x = value))+
    geom_histogram(fill=NA, colour = "black")+
    facet_wrap(.~name, scales = "free")
```

Now we move forward to using using the `brms` package to calculate the posterior of the normal distribution

```{r estimate_posterior_brms, message=FALSE, warning=FALSE}


hd_adults_height <- hd_adults |> select(height) 

height_post <- (
  hd_adults_height |> 
    brm(
      data = _, 
      # since there is only intercept, this will estimate the average height 
      formula = height ~ 1, 
      family = gaussian, 
      prior = c(
        prior(normal(178, 28), class = Intercept),
        prior(cauchy(5, 1), class = sigma)
      ), 
      iter = 2000, warmup = 1000, chains = 4, cores = 4, seed = 72, 
      file = "../fits/height_post"
  )
)

# check some prelimnary diagnostics
plot(height_post) 

# generates a dataframe of poterior samples 
height_post_df <- (
  height_post |> 
    posterior::as_draws_df() 
    )

height_post_cov <- height_post_df |> select(b_Intercept:sigma) |> cov()

# some ways of summarizing the information of the posterior distribution 
# 1. These generate the quantiles of the posterior distribution.
posterior_summary(height_post, probs = c(0.025, 0.11, 0.5, 0.89, 0.975))

# 2. These generate the x% higheset density intervals. defaiult is 95%
height_post_df |> 
  pivot_longer(cols = b_Intercept:lp__) |>
  group_by(name) |> 
  mean_hdi()
  

# chain plot
# the extracted posterior distribution can be used to generate custom diagnotic plots 
# eg. the chains of the posterior distribution
height_post_df |> 
  select(-c(Intercept, lprior, lp__)) |> 
  pivot_longer(cols = b_Intercept:sigma, names_to = "param_name", 
               values_to = "param_value") |> 
  ggplot(aes(x = .iteration, y = param_value, colour = factor(.chain))) +
  geom_line(size = 0.4)+
  facet_wrap(.~param_name, scales = "free_y")
    

# eg. poesterior distribution
height_post_df |> 
  select(-c(Intercept, lprior, lp__)) |> 
  pivot_longer(cols = b_Intercept:sigma, 
               names_to = "param_name", values_to = "param_value") |> 
  ggplot(aes(x = param_value)) +
  geom_histogram(bins = 50, fill = NA, colour = "black")+
  facet_wrap(.~param_name, scales = "free_x")

```

Similarly, moving on to form a linear model between height and weight

```{r linear_mod_ht_wt, warning=FALSE, message=FALSE}

# extract data 
hd_adults_ht_wt <- (
  hd_adults |> 
    select(height, weight) 
  )
  

# prior with data
n_lines <- 500

prior_lines <- (
  tibble(
    line_id = 1:n_lines, 
    a = rnorm(n_lines, mean = 178, sd = 20), 
    # since otherwise the relationships are going to look absurd otherwise 
    b = rlnorm(n_lines, mean = 0, sd = 1)
    ) |> 
    expand_grid(weight = range(hd_adults_ht_wt$weight)) |> 
    mutate(height_pred = a + b*(weight-mean(weight)))
  )


# make a tibble to annotate the plot
text <-
  tibble(weight = c(33, 36),
         height_pred = c(0 - 25, 272 + 25),
         label  = c("Embryo", "World's tallest person (272 cm)"))


# plotting these lines
prior_lines |> 
  ggplot(aes(x = weight, y = height_pred)) +
  geom_hline(yintercept = c(0, 272), linetype = 2:1, linewidth = 1/3) +
  geom_line(aes(group = line_id), alpha = 1/10) +
  geom_text(data = text,
            aes(label = label),
            size = 3) +
  coord_cartesian(ylim = c(-100, 400)) 
  
  
  
  

```

Using `brms` to estimate the posterior distribution of this linear relationship

```{r line_post_distn, warning = FALSE, message = FALSE}

# this data is going to be used estimating the posterior distribution 
hd_adults_ht_wt_2 <- (
    hd_adults_ht_wt |> 
    # recenter weights at the average in order to make 
    # comparsions across multiple populations 
    mutate(weight_c = weight - mean(weight))
  )


# calculate the posterior using the priors above 
ht_wt_post <- (
  hd_adults_ht_wt_2 |> 
    brm(
      formula = height ~ 1 + weight_c, 
      family = gaussian, 
      prior = c(
        prior(normal(178, 28), class = Intercept),
        prior(lognormal(0,1), class = b),
        prior(uniform(0, 50), class = sigma, lb = 0, ub = 50)
      ), 
      iter = 2000, warmup = 1000, chains = 4, cores = 4, seed = 72, 
      file = "../fits/height_weight_post"
    )
  )

summary(ht_wt_post)

plot(ht_wt_post)
# similarly fitting a different formulation -- trying out the syntax of specifying formulae in brms

# calculate the posterior using the priors above 
ht_wt_post_nonlin <- (
  hd_adults_ht_wt_2 |> 
    brm(
      formula = bf(height ~ a + exp(lb)*weight_c,
        a ~ 1, 
        lb ~ 1, 
        nl = TRUE
        ), 
      family = gaussian, 
      prior = c(
        prior(normal(178, 28), class = b, nlpar = a),
        prior(normal(0,1), class = b, nlpar = lb),
        prior(cauchy(5, 1), class = sigma)
      ), 
      iter = 2000, warmup = 1000, chains = 4, cores = 4, seed = 72, 
      file = "../fits/height_weight_nonlin_post"
    )
  )

summary(ht_wt_post_nonlin)

plot(ht_wt_post_nonlin)

# To recover the values from the prveious linear fit, need to antilog the estimate
fixef(ht_wt_post_nonlin)["lb_Intercept", "Estimate"] |> exp() |> round(3)
# compared to the slope estimate from earlier
fixef(ht_wt_post)["weight_c", "Estimate"] |> round(3)


# Viewing the tables of marginal distributions
posterior_summary(ht_wt_post)[1:3,] |> round(3)


# calculating the variance-covariance among the parameters 
ht_wt_post |> 
  as_draws_df() |> 
  select(b_Intercept, b_weight_c, sigma) |> 
  cov() |> 
  round(3)

# visualizing the variance covariance matrix
ht_wt_post |> 
  as_draws_df() |> 
  select(b_Intercept, b_weight_c, sigma) |> 
  ggpairs(
    diag = list(continuous = wrap("barDiag", bins = 50, colour = "black", fill = NA)), 
    lower = list(continuous = wrap("points", shape = 21, 
                                   colour = "black",
                                   fill = "blue", alpha = 0.6)
                 )
  ) 




```

### Posterior Predictive Check –

After having convinced yourself about the estimate of the posterior distribution. It is finally time compare the predictions from the estimated model with the data. Its is here that we also visualize how uncertainty plays into capturing the variation in the data.

```{r post_pred_checks, warning = FALSE, message = FALSE}

# note that the model was fit to a recentered distribution of weights. To represent 
# the original values, the follwoing plot will be relabeled

labels <- (
  c(-10, 0, 10) + mean(hd_adults_ht_wt$weight) |> 
    round(digits = 0)
  )


# fitting the mean line through the data 
hd_adults_ht_wt_2 |> 
  ggplot(aes(x = weight_c, y = height)) +
  geom_point(shape = 21) +
  geom_abline(
    intercept = fixef(ht_wt_post)['Intercept', 'Estimate'],
    slope     = fixef(ht_wt_post)['weight_c', 'Estimate'],
    colour = "#D55E00"
    )+
  scale_x_continuous("weight",
                     breaks = c(-10, 0, 10),
                     labels = labels) 

# demonstrating uncertainty around the regression lines 
ht_wt_post_df <- (
  ht_wt_post |> 
    as_draws_df() 
  )


hd_adults_ht_wt_2 |> 
  ggplot(aes(x = weight_c, y = height)) +
  geom_point(shape = 21) +
  geom_abline(
    data = ht_wt_post_df |> 
      slice(1:30), 
    aes(
      intercept = b_Intercept,
      slope     = b_weight_c),
    colour = "#D55E00", 
    size = 0.1
    )+
  scale_x_continuous("weight",
                     breaks = c(-10, 0, 10),
                     labels = labels) 

```

That is all cool. But the main object of fitting a model is to generate prediction at particular values of the predictor. Lets say that we are interested in predicting the values of height for the 50kg interval.

```{r regression_intervals}

# first generating a prediction distribution at the average value 
mean_wt <- hd_adults_ht_wt_2 |> summarise_all(mean) |> select(weight) |> unlist()

# then
mu_at_50_dist <- (
  ht_wt_post_df |> 
    transmute(mu = b_Intercept + (50-mean_wt)*b_weight_c)
) 
  
# generating the table of uncertainty at 50kg weight
mean_hdi(mu_at_50_dist, .width = c(0.89, 0.95))


# corresponding plot showing the distribution of mu 
mu_at_50_dist |> 
  ggplot(aes(x = mu)) +
  stat_halfeye(point_interval = mode_hdi, .width = .95,
               fill = '#1f77b4') +
  #scale_y_continuous(NULL, breaks = NULL) +
  labs(
    x = expression(mu["height | weight = 50"]), 
    y = "probability"
    ) 
  
# similalry using brms::fitted, one can calulate average predictions for a arbitrary
# data 

# lets use values from the exrecise problems 
weight_seq <- (
  tibble(
    weight = c(46.95, 43.72, 64.78, 32.59, 54.63)
  ) |>
    # recenter this new data according to the mean wt from the previous data to which
    # the model was fit
    mutate(
      individual = 1:5,
      weight_c = weight - mean_wt
      )
)
  
# summarised prediction for mu at different are therefore given as 
pred_mu_arb_sum <- (
  weight_seq |> 
    bind_cols(
      fitted(
        object = ht_wt_post, 
        newdata = weight_seq, 
        summary = TRUE, 
        probs = c(0.105, 0.785)
        ) |> 
        as_tibble()    
    )
  
)
  
# now generating the prediction for all values of weight in he original dataset
pred_mu_df <- (
  fitted(object = ht_wt_post, 
         newdata = hd_adults_ht_wt_2, summary = FALSE) |>  
  as_tibble() |> 
  rlang::set_names(1:nrow(hd_adults_ht_wt_2)) |> 
  mutate(iter = 1:n()) |> 
  pivot_longer(cols = -iter, 
               names_to = "id", 
               values_to = "pred_mu"
               )
  
)

hd_adults_wt_2 <- hd_adults_ht_wt_2 |> 
  mutate(id = 1:n() |>as.character()) |> 
  select(id, weight, height)



pred_mu_plot_df <- (
  pred_mu_df |>  
    right_join(hd_adults_wt_2, by = "id")
    )

pred_mu_plot_df |> 
  group_by(id) |> 
  sample_n(50, replace = FALSE) |> 
  ungroup() |> 
  ggplot(aes(x = weight, y = pred_mu, group = iter)) +
  geom_point(shape = 21, colour = '#1f77b4', fill = NA)+
  labs(y = "height")


# finally the prediction intervals 
# first: summarise posterior predicted average mu 
pred_mu_df_sum <- (
  fitted(object = ht_wt_post, 
         newdata = hd_adults_ht_wt_2,  probs = c(0.025, 0.975)) |> 
  as_tibble() |> 
  mutate(id = 1:n() |> as.character()) |> 
  right_join(
      hd_adults_wt_2, by = "id"
    )
  
)

# second: summarise the posterior predictive distribution of the response
pred_height_df_sum <- (
  predict(object = ht_wt_post, 
         newdata = hd_adults_ht_wt_2,  probs = c(0.025, 0.105, 0.975, 0.785)) |> 
  as_tibble() |> 
  mutate(id = 1:n() |> as.character()) |> 
  right_join(
      hd_adults_wt_2, by = "id"
    )
  
)


# plot the posterior preditive distribution of the rsponse and the average.
pred_mu_df_sum |> 
  ggplot(aes(x = weight))+
  geom_point(aes(y = height), shape = 21, colour = "black", fill = NA) +
  geom_ribbon(data = pred_height_df_sum, 
              aes(ymin = Q2.5, ymax = Q97.5), 
              fill = "#D55E00", alpha = 0.3
              ) +
  geom_ribbon(data = pred_height_df_sum, 
              aes(ymin = Q10.5, ymax = Q78.5), 
              fill = "#D55E00", alpha = 0.5
              ) +
  geom_smooth(aes(y = Estimate, ymin = Q2.5, ymax = Q97.5), 
              colour = "#F0E442", fill = "#D55E00", stat = "identity", alpha = 0.8) 




```

### Exercises

Carrying out some rudimentary analyses on the rest of the data.

This is an exercise set from the **statistical rethinking** textbook. Starting out with the Howell data concentrating only on individuals who are younger than 18 years

```{r exercises_1, message=FALSE, warning=FALSE}
# selecting out individuals 

# subset the Howell data to include only adults 
hd_young <- hd |> filter(age < 18) 

# a quick plot to look at the relationship between height and weight
hd_young |> 
  ggplot(aes(x = weight, y = height))+
  geom_point(pch = 21, fill = NA, , colour = "black")


# fitting a linear model to this data 
# center the weights around the mean 
hd_young_2 <- hd_young |> mutate(weight_c = weight - mean(weight))

ht_wt_young_post <- (
  hd_young_2 |> 
    brm(
      data = _,
      formula = height ~ 1 + weight_c, 
      family = gaussian, 
      prior = c(
        prior(normal(178, 28), class = Intercept),
        prior(lognormal(0,1), class = b),
        prior(inv_gamma(1, 0.5), class = sigma)
      ), 
      iter = 2000, warmup = 1000, chains = 4, cores = 4, seed = 72, 
      file = "../fits/ht_wt_young_post"
    )
  )

# how does the summary of estimates look like 

summary(ht_wt_young_post)

# Based on the marginal distribution for parameter estimates every 10 kg increase in weight, 
# the height shows a increase of 27.2 cm 

# the prediction intervals 
# first: summarise predicted mu 

pred_mu_young_df_sum <- (
  fitted(object = ht_wt_young_post, 
         newdata = hd_young_2,  probs = c(0.025, 0.975)) |> 
  as_tibble() |> 
  mutate(id = 1:n() |> as.character()) |> 
  right_join(
      hd_young_2 |> 
        mutate(id = 1:n() |> as.character()), 
      by = "id"
    )
)


pred_height_young_df_sum <- (
  predict(object = ht_wt_young_post, 
         newdata = hd_young_2,  probs = c(0.025, 0.105, 0.975, 0.785)) |> 
  as_tibble() |> 
  mutate(id = 1:n() |> as.character()) |> 
  right_join(
      hd_young_2 |> 
        mutate(id = 1:n() |> as.character()), 
      by = "id"
    )
  
)


# plot the posterior preditive distribution 
pred_mu_young_df_sum |> 
  ggplot(aes(x = weight))+
  geom_point(aes(y = height), shape = 21, colour = "white", fill = NA) +
  geom_ribbon(data = pred_height_young_df_sum, 
              aes(ymin = Q2.5, ymax = Q97.5), 
              fill = "#D55E00", alpha = 0.3
              ) +
  geom_ribbon(data = pred_height_young_df_sum, 
              aes(ymin = Q10.5, ymax = Q78.5), 
              fill = "#D55E00", alpha = 0.5
              ) +
  geom_smooth(aes(y = Estimate, ymin = Q2.5, ymax = Q97.5), 
              colour = "#F0E442", fill = "#D55E00", stat = "identity", alpha = 0.8) 
```