---
title: "Hidden markov models"
author: "Deven Gokhale"
format: 
  html:
    theme: sketchy
    code-fold: true
editor: visual
---

```{r load_libraries}
#| warning: false
#| message: false
rm(list = ls())
set.seed(42)

# packages ---
# system operations
library(here)
library(fs)

# inference and simulation ---
library(cmdstanr)
# setting-up stan
cmdstan_path <- "C:/Users/gokhad01/.cmdstan/cmdstan-2.37.0"
torsten_path <- "C:/Users/gokhad01/.torsten/cmdstan"

set_cmdstan_path(path = cmdstan_path)
# set up directories to store stan files

library(brms)
library(posterior)
library(tidybayes)
library(mrgsolve)

# data wrangling
library(dplyr)
library(extraDistr, include.only = c("dhcauchy"))
library(psych, include.only = "describe")
library(tidyr, include.only = c("pivot_longer", "pivot_wider"))
library(tibble, include.only = "tibble")
library(purrr)

# plotting
library(ggplot2)
library(ggdist)
library(patchwork)
library(dagitty)
library(ggdag)
library(scico)
library(assertthat)


# set up tghe plotting theme
theme_set(theme_bw(base_size = 15))

# establish the project paths
here(".")
here::i_am(
  "./notebooks/hidden_markov_models.qmd"
)
```


```{css notebook_appearance, echo: false}
.cell-output-stdout {
  max-height: 500px;  
  overflow-y: auto;
}
```

```{r, utility_functions}

# this function generates transition probability for 1 --> 2 given some covariates.
cal_p12 <- function(Cavg, beta0 = -5.1, beta1 = 1.0) {
  logit_p12 <- beta0 + beta1 * Cavg
  return(inv_logit_scaled(logit_p12))
}

# this function will assign the
get_transition <- function(p12) {
  # check the input data
  assert_that(is.numeric(p12))
  assert_that(p12 <= 1)

  transition_matrix <- matrix(
    c((1 - p12), p12, 0.05, 0.95),
    byrow = TRUE,
    nrow = K
  )
  return(transition_matrix)
}

# this function produces the panel of the hidden states and emmissions together with
# the behavior of the dependent variable.
plot_sim_panels <- function(sim_data) {
  # browser()
  # check that data before using it
  assert_that(is.data.frame(sim_data))
  assert_that(
    sum(
      c("times", "latent_state", "adverse_events", "Cavg") %in%
        names(sim_data)
    ) ==
      4
  )

  # plot the emissions
  emissions_plot <- sim_data |>
    ggplot(aes(
      x = times,
      y = adverse_events,
      colour = latent_state
    )) +
    geom_point(size = 3) +
    labs(
      x = "Time (Days)",
      y = "Adverse Events",
      colour = "Latent\nState"
    )

  # plot hidden states as a function of the time varying covariate
  hidden_states_plot <- sim_data |>
    mutate(
      state_numeric = as.integer(latent_state),
      rolling_state = slider::slide_dbl(
        state_numeric,
        mean,
        .before = 10,
        .after = 10
      )
    ) |>
    ggplot(aes(x = times)) +
    geom_step(
      aes(y = state_numeric, colour = "Latent state"),
      linewidth = 0.4,
      alpha = 0.6
    ) +
    geom_line(aes(y = rolling_state, colour = "Rolling mean"), linewidth = 1) +
    geom_line(
      aes(
        y = (Cavg - min(Cavg)) / (max(Cavg) - min(Cavg)) + 1,
        colour = "Cavg"
      ),
      linetype = "dashed",
      linewidth = 1
    ) +
    scale_colour_manual(
      name = NULL,
      values = c(
        "Latent state" = "grey70",
        "Rolling mean" = "steelblue",
        "Cavg" = "firebrick"
      )
    ) +
    scale_y_continuous(
      name = "Latent state / rolling mean",
      limits = c(1, 2),
      sec.axis = sec_axis(
        transform = ~ (. - 1) * (max(Cavg) - min(Cavg)) + min(Cavg),
        name = "Cavg"
      )
    ) +
    labs(x = "Time (Days)") +
    theme(legend.position = "bottom")

  # putting it all together
  emissions_plot /
    hidden_states_plot +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom") &
    guides(colour = guide_legend(nrow = 2))
}
```
### Simulating a Hidden markov model

# Simulation 

I am going to write models that that will become progressively more complex. Here I am trying to model a relationship between exposure of a drug and the incendence of adverse events. 

### Quick summary of the model 

Here, I assume a two, hidden states model that scafolds the risk structure associated with being "exposed". Lets say **state 1** is the "low risk" group and **state 2** is the "high risk" group. The main difference here being that the high risk group emits appreciably higher average number of the "adeverse events" than the lower risk group. The over all dynamic structire of this model is decribed in the flow diagram below. 

<!-- Insert diagram here --> 




## Constant transition probabilities 

Trying to play around some of the models that center around simulating adverse events -- I am assuming the probabilities of transition don't change... However, I assume that that once you are in state 2, it becomes becomes diffcult to get out of it.  


```{r, hmm_first_pass}
# setting up constants of integration
T <- 365
K <- 2

# times for later
times <- 1:T

# currently assuming a very simple transition matrix
transition_matrix <- matrix(
  c(0.2, 0.8, 0.3, 0.7),
  byrow = TRUE,
  nrow = K
)

# average ae incidence
lambda <- c(2.1, 9.7)
# initializing the system
current_state <- c(1, rep(NA, times = T - 1))
emissions <- rep(NA, times = T)

# simulate the hmm
for (t in 2:T) {
  #  sample from the latent state distribution
  # browser()
  current_state[t] <- sample(
    x = 1:K,
    size = 1,
    prob = transition_matrix[current_state[(t - 1)], ],
    replace = TRUE
  )
  #  generate emissions
  emissions[t] <- rpois(1, lambda[current_state[t]])
}

plot_simulations <- tibble(
  times = times,
  latent_state = as.factor(current_state),
  adverse_events = emissions
)

# plotting the process
plot_simulations |>
  ggplot(aes(
    x = times,
    y = adverse_events,
    colour = latent_state
  )) +
  geom_point(size = 3) +
  labs(
    x = "Time (Days)",
    y = "Adverse Events",
    colour = "Latent\nState"
  ) +
  theme(legend.position = "bottom")
```


## Time varying transition probabilities 

### $C_{\text{avg}}$ has temporal dynamics
#### Cosine 
```{r hmm_forced_transition_sine}
#| fig-height: 8
#| fig-width: 9

# setting up constants of integration
T <- 365
K <- 2

# times for later
times <- 1:T


# introducing time varying effects
Cavg <- (1.5 - cos(2 * pi * times / T))

p12 <- cal_p12(Cavg)

# average ae incidence
lambda <- c(2.1, 9.7)
# initializing the system
current_state <- c(1, rep(NA, times = T - 1))
emissions <- rep(NA, times = T)


# simulate the hmm
for (t in 2:T) {
  #  sample from the latent state distribution
  # browser()
  # current transition matrix
  current_transition_matrix <- get_transition(p12[t - 1])
  current_state[t] <- sample(
    x = 1:K,
    size = 1,
    prob = current_transition_matrix[current_state[(t - 1)], ],
    replace = TRUE
  )
  #  generate emissions
  emissions[t] <- rpois(1, lambda[current_state[t]])
}


plot_simulations <- tibble(
  times = times,
  latent_state = as.factor(current_state),
  adverse_events = emissions,
  Cavg = Cavg
)


plot_sim_panels(plot_simulations)
```

#### 2 compartment PK-model

I am goint to simulate from a 2cmt PK model with linear absorption. 
<!-- Insert figure here -->

```{r pk_model}
#| warning: false
#| message: false
#| fig-width: 8
#| fig-cap: "Two-Compartment PK: 50 mg \u2192 100 mg at Week 24 (Every 4 Weeks Dosing)"

# define two-compartment model
code <- '
$PARAM CL = 0.3, V1 = 10, Q = 2, V2 = 25, KA = 0.4

$CMT DEPOT CENTRAL PERIPH

$ODE
dxdt_DEPOT   = -KA * DEPOT;
dxdt_CENTRAL =  KA * DEPOT - (CL/V1 + Q/V1) * CENTRAL + (Q/V2) * PERIPH;
dxdt_PERIPH  =  (Q/V1) * CENTRAL - (Q/V2) * PERIPH;

$TABLE
double CP = CENTRAL / V1;

$CAPTURE CP
'

mod <- mcode("two_cmt", code)

# dosing events: weekly (ii = 168 hrs)
# weeks 1-23: 50 mg (23 doses), weeks 24-52: 100 mg (29 doses)
doses_low <- ev(amt = 50, ii = 168, addl = 22, time = 0, cmt = 1)
doses_high <- ev(amt = 100, ii = 168, addl = 28, time = 23 * 168, cmt = 1)

dosing <- c(doses_low, doses_high)

# simulate over 52 weeks with hourly output
sim_pk <- mod |>
  ev(dosing) |>
  mrgsim(end = 52 * 168, delta = 1) |>
  as_tibble() |>
  mutate(day = floor(time / 24)) |>
  mutate(Cavg = mean(CP), .by = day)

# produce the simulation
head(sim_pk)

# plot daily Cavg
sim_pk |>
  ggplot(aes(x = time / 24, y = Cavg)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_vline(
    xintercept = 168,
    linetype = "dashed",
    colour = "firebrick",
    linewidth = 0.7
  ) +
  annotate(
    "text",
    x = 168,
    y = max(sim_pk$Cavg) * 0.9,
    label = "Dose escalation\n50 \u2192 100 mg",
    hjust = 0,
    colour = "firebrick",
    size = 3.5
  ) +
  labs(
    x = "Time (days)",
    y = "Daily Cavg (mg/L)",
    title = "Two-Compartment PK: Daily Cavg, 50 mg \u2192 100 mg at Week 24 (Weekly Dosing)"
  )
```

Now resimulate the satates and emmsions with this new $C_{\text{avg}}$ time-series.

```{r hmm_cmt_pk_forcing}
#| fig-height: 8
#| fig-width: 9

# extract one daily Cavg from the PK sim (days 1-365)
pk_daily <- sim_pk |>
  distinct(day, Cavg) |>
  filter(day >= 0, day <= T - 1) |>
  arrange(day)

Cavg <- pk_daily$Cavg

p12 <- cal_p12(Cavg)

# average ae incidence (same as cosine scenario)
lambda <- c(2.1, 9.7)

# initialise
current_state <- c(1, rep(NA, times = T - 1))
emissions <- rep(NA, times = T)

# simulate the hmm
for (t in 2:T) {
  current_transition_matrix <- get_transition(p12[t - 1])
  current_state[t] <- sample(
    x = 1:K,
    size = 1,
    prob = current_transition_matrix[current_state[(t - 1)], ],
    replace = TRUE
  )
  emissions[t] <- rpois(1, lambda[current_state[t]])
}

plot_simulations <- tibble(
  times = times,
  latent_state = as.factor(current_state),
  adverse_events = emissions,
  Cavg = Cavg
)

plot_sim_panels(plot_simulations)
```