---
title: "Meta-analysis from Solomon Kurz's blog"
author: "Deven Gokhale"
format: 
  html:
    theme: sketchy
    code-fold: true
editor: visual
---

```{css notebook_appearance, echo: false}
.cell-output-stdout {
  max-height: 500px;  
  overflow-y: auto;
}
```

```{r}
#| warning: false
#| message: false
rm(list = ls())

# packages ---
# system operations
library(here)
library(fs)

# inference ---
library(cmdstanr)
# setting-up stan
cmdstan_path <- "C:/Users/gokhad01/.cmdstan/cmdstan-2.37.0"
torsten_path <- "C:/Users/gokhad01/.torsten/cmdstan"

set_cmdstan_path(path = cmdstan_path)
# set up directories to store stan files

library(brms)
library(posterior)
library(tidybayes)

# data wrangling
library(dplyr)
library(extraDistr, include.only = c("dhcauchy"))
library(psych, include.only = "describe")
library(tidyr, include.only = c("pivot_longer", "pivot_wider"))
library(tibble, include.only = "tibble")
library(purrr)

# plotting
library(ggplot2)
library(ggdist)
library(patchwork)
library(dagitty)
library(ggdag)
library(scico)


# set up tghe plotting theme
theme_set(theme_bw(base_size = 15))

# establish the project paths
here(".")
here::i_am(
  "./notebooks/meta_analysis.qmd"
)
```

# Cats adoption

This is example from statistical rethinking walks through incorporating the censorship model in a survival analysis example.

## Research question

The "research" question at hand is does the population of "black" cats have a higher waiting time time to adoption relative to cat population with other fur colours

### Simulating TTE for cats -- 

This set of functions show how to simulate the time to an event, in this case "Adoption" of a cat of certain color. Theoretically, this event of the 1 succsess follwos a geormtric distribution whos pdf is given by the following expression -- $P(D_i|A = 1) = (1-p_C)^{(D_i - 1)}\times p_c$

The actual generation of this geometric distribution can be realized as follows-- 

```{r sim_tte_from_scratch}
# function to calculate the days it takes to adopt a cat.
cat_adopt <- function(day, prob) {
  if (day > 1000) {
    return(day)
  }

  if (runif(1) > prob) {
    # keep waiting
    day <- cat_adopt(day + 1, prob)
  }

  return(day)
}

# simulate days until adoption --> There is no censoring in this data generation process.
sim_cats1 <- function(n = 10, probs = c(0.1, 0.2)) {
  colour <- rep(NA, n)
  days <- rep(NA, n)

  for (i in 1:n) {
    colour[i] <- sample(c(1, 2), size = 1, replace = TRUE)
    days[i] <- cat_adopt(1, probs[colour[i]])
  }

  return(
    tibble(
      N = n,
      days = days,
      colour = colour,
      adopted = rep(1, n)
    )
  )
}

set.seed(42)
synth_cats <- sim_cats1(n = 1000)

day0_pad <-
  tibble(
    N = 1000,
    days = 0,
    colour = c(1, 2),
    total_adopted = 0,
    prop_unadopted = 1
  )

synth_cats_data <- synth_cats |>
  group_by(colour) |>
  arrange(days) |>
  mutate(
    total_adopted = cumsum(adopted),
    prop_unadopted = (n() - total_adopted) / n()
  ) |>
  ungroup() |>
  bind_rows(day0_pad) |>
  arrange(colour, days)


# quick plot to visualize the survivorship
synth_cats_data |>
  ggplot(aes(x = days, y = prop_unadopted, colour = factor(colour))) +
  geom_step(linewidth = 0.9) +
  scale_colour_manual(
    values = c("1" = "black", "2" = "steelblue"),
    labels = c("1" = "black", "2" = "other")
  ) +
  labs(
    x = "Time (Days)",
    y = "Proportion un-adopted",
    colour = "Fur colour",
  ) +
  scale_x_continuous(lim = c(0, 50), breaks = seq(0, 50, 10))
```

### Simulating TTE for cats

I am going to redo the step above using the `rgeom()` function in R. From the example above there are 517 black and 485 white cats in the generated sample above. I will keep that structure intact to overly the simulation realated variance on top of the synthetic survivorship of the population above.  

```{r sim_tte_geometric}

tte_cat_adop <- function(n = 517, prob = 0.1, colour = 1) {
  # pad the result for better representation
  # browser()
  day0_pad <-
    tibble(
      N = n,
      days = 0,
      colour = colour,
      total_adopted = 0,
      prop_unadopted = 1
    )

  unadopted_survivorship <- tibble(
    N = n,
    days = rgeom(n, prob) + 1,
    adopted = 1L,
    colour = colour
  ) |>
    mutate(
      total_adopted = cumsum(adopted),
      prop_unadopted = (n - total_adopted) / n
    ) |>
    bind_rows(day0_pad) |>
    arrange(days)
  return(unadopted_survivorship)
}

sim_cats <- function(make_up_data) {
  map_dfr(1:nrow(make_up_data), \(x) {
    n <- make_up_data |> slice(x) |> pull(n)
    prob <- make_up_data |> slice(x) |> pull(prob)
    colour <- make_up_data |> slice(x) |> pull(colour)

    return(tte_cat_adop(n, prob, colour))
  })
}

tibble(
  n = c(517, 485),
  prob = c(0.1, 0.2),
  colour = c(1, 2)
) |>
  sim_cats()
```

### Rumblings with Stan

There is a type of syntax that I have seen pop up especially in some more advnaced models. I want to figure what these code do. 

```{r stan_code_tutorial}

b <- c(2, 0.75)

# simulate some
synth_data <- tibble(
  x = rnorm(1e2),
  y = rnorm(1e2, b[1] * exp(b[2] * x), 2.5)
)

synth_data |>
  ggplot(aes(x, y)) +
  geom_point(shape = 21)


# priors
priors <- (prior(normal(1, 2), nlpar = "b1") +
  prior(normal(0, 2), nlpar = "b2") +
  prior(exponential(0.1), class = "sigma"))


# bayesain formula
bf <- bf(y ~ b1 * exp(b2 * x), b1 + b2 ~ 1, nl = TRUE)


# describing the model in terms of the brm formalizm
mod1 <- brm(
  formula = bf,
  prior = priors,
  data = synth_data,
  family = gaussian(),
  cores = 4,
  seed = 42,
  backend = "cmdstanr"
)

# what do the posterior as ---

summary(mod1)

pp_check(mod1)

# posterior predictive checks
mod1_post_df <- as_draws_df(mod1)

x_grid <- seq(min(synth_data$x), max(synth_data$x), length.out = 100)

curve_df <- mod1_post_df |>
  slice(1:30) |>
  mutate(.draw = row_number()) |>
  reframe(
    x = x_grid,
    y_hat = b_b1_Intercept * exp(b_b2_Intercept * x),
    .by = .draw
  )

synth_data |>
  ggplot(aes(x, y)) +
  geom_point(shape = 21) +
  geom_line(
    data = curve_df,
    aes(x = x, y = y_hat, group = .draw),
    colour = "#D55E00",
    linewidth = 0.1
  ) +
  labs(x = "x", y = "y", title = "Posterior predictive curves (30 draws)")


# want to see how the stan code looks based on the brms pacakge --
stancode(mod1)


# look  at the draws dataframe
mod1 |> print()

posterior_summary(mod1)

as_draws_df(mod1) |> glimpse()
```



```{r, rewriting_model_in_stan_v1}

# using compose data
stan_data <- synth_data |>
  compose_data()

# compile the stan model
cmd_mod11 <- cmdstan_model("notebooks/mod11.stan")

cmd_mod11_fit <- cmd_mod11$sample(
  data = stan_data,
  chains = 4,
  seed = 42
)

print(cmd_mod11_fit)
```

```{r, rewriting_model_in_stan_v2}

# compile the stan model
cmd_mod12 <- cmdstan_model("notebooks/mod12.stan")

cmd_mod12_fit <- cmd_mod12$sample(
  data = stan_data,
  chains = 4,
  seed = 42
)

print(cmd_mod12_fit)
```


### Mixture Beta distribution 

#### Generate data 

```{r sim_beta_mixture}

n <- 1000

sim_mix_beta <- function(
  n,
  prob = c(0.1, 0.8, 0.1),
  alpha = 1,
  beta = 5
) {
  # browser()
  # generate distribution of the componenent
  component <- sample(x = c(1, 2, 3), size = n, replace = TRUE, prob = prob)
  beta_draw <- rbeta(n, alpha, beta)
  # simulate from the mixture model
  mixture_sample <- case_when(
    component == 1 ~ 0,
    component == 3 ~ 1,
    TRUE ~ beta_draw
  )

  return(list(component, mixture_sample))
}

set.seed(42)
sim_mix_beta(100)

# tibble(
#   x = rnorm(n)
# )
```



### Simulating a HMM

#### Constant transition probabilities 

Trying to play around some of the models that center around simulating adverse events -- I am assuming the probabilities of transition don't change... However, I assume that that once you are in state 2, it becomes becomes diffcult to get out of it.  


```{r, hmm_first_pass}
# setting up constants of integration
T <- 365
K <- 2

# times for later
times <- 1:T

# currently assuming a very simple transition matrix
transition_matrix <- matrix(
  c(0.2, 0.8, 0.3, 0.7),
  byrow = TRUE,
  nrow = K
)

# average ae incidence
lambda <- c(2.1, 9.7)
# initializing the system
current_state <- c(1, rep(NA, times = T - 1))
emissions <- rep(NA, times = T)

# simulate the hmm
for (t in 2:T) {
  #  sample from the latent state distribution
  # browser()
  current_state[t] <- sample(
    x = 1:K,
    size = 1,
    prob = transition_matrix[current_state[(t - 1)], ],
    replace = TRUE
  )
  #  generate emissions
  emissions[t] <- rpois(1, lambda[current_state[t]])
}

plot_simulations <- tibble(
  times = times,
  latent_state = as.factor(current_state),
  adverse_events = emissions
)

# plotting the process
plot_simulations |>
  ggplot(aes(
    x = times,
    y = adverse_events,
    colour = latent_state
  )) +
  geom_point(size = 3) +
  labs(
    x = "Time (Days)",
    y = "Adverse Events",
    colour = "Latent\nState"
  ) +
  theme(legend.position = "bottom")
```


#### Time varying transition probabilities 


```{r hmm_forced_transition}

# setting up constants of integration
T <- 365
K <- 2

# times for later
times <- 1:T

# this function will assign the
get_transition <- function(p12) {
  transition_matrix <- matrix(
    c((1 - p12), p12, 0.05, 0.95),
    byrow = TRUE,
    nrow = K
  )
  return(transition_matrix)
}

# introducing time varying effects
Cavg <- (1.5 - cos(2 * pi * times / T))
# generate the time series of transition probabilities
# beta0=-3.5, beta1=1.0 gives:
#   trough (Cavg~0.5): p12 ≈ 0.05 → mean sojourn state 1 ~21 days
#   peak   (Cavg~2.5): p12 ≈ 0.27 → mean sojourn state 1 ~4 days
cal_p12 <- function(Cavg, beta0 = -3.5, beta1 = 1.0) {
  logit_p12 <- beta0 + beta1 * Cavg
  return(inv_logit_scaled(logit_p12))
}

p12 <- cal_p12(Cavg)

# average ae incidence
lambda <- c(2.1, 9.7)
# initializing the system
current_state <- c(1, rep(NA, times = T - 1))
emissions <- rep(NA, times = T)


# simulate the hmm
for (t in 2:T) {
  #  sample from the latent state distribution
  # browser()
  # current transition matrix
  current_transition_matrix <- get_transition(p12[t - 1])
  current_state[t] <- sample(
    x = 1:K,
    size = 1,
    prob = current_transition_matrix[current_state[(t - 1)], ],
    replace = TRUE
  )
  #  generate emissions
  emissions[t] <- rpois(1, lambda[current_state[t]])
}


plot_simulations <- tibble(
  times = times,
  latent_state = as.factor(current_state),
  adverse_events = emissions
)

# plotting the process
plot_simulations |>
  ggplot(aes(
    x = times,
    y = adverse_events,
    colour = latent_state
  )) +
  geom_point(size = 3) +
  labs(
    x = "Time (Days)",
    y = "Adverse Events",
    colour = "Latent\nState"
  ) +
  theme(legend.position = "bottom")

# rolling mean state vs rescaled Cavg
plot_simulations |>
  mutate(
    state_numeric = as.integer(latent_state),
    rolling_state = slider::slide_dbl(
      state_numeric,
      mean,
      .before = 14,
      .after = 14
    )
  ) |>
  ggplot(aes(x = times)) +
  geom_step(
    aes(y = state_numeric),
    colour = "grey70",
    linewidth = 0.4,
    alpha = 0.6
  ) +
  geom_line(aes(y = rolling_state), colour = "steelblue", linewidth = 1) +
  geom_line(
    aes(y = scales::rescale(Cavg, to = c(1, 2))),
    colour = "firebrick",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    x = "Time (Days)",
    y = "State / Rolling mean (blue) / rescaled Cavg (red)"
  )
```